Make any stupid plot you want

### 2021-05-09 draft/sections/implementation.tex
clean ups

      def draw(self, renderer):
          # returns K, indexed on fiber then key 
          tau = self.E.view(self.axes) 
          # visual channel encoding applied fiberwise 
          mu = {p: nu(tau(c))
                   for p, (c, nu) in self.V.items()} 
          self.q_hat(**mu)
          # pass configurations off to the renderer
          super().draw(renderer)
  \end{minted}
 The data is fetched in section \dsection\ via a \mintinline{python}{view} method on the data because the input to the artist is a section on \dtotal. The \mintinline{python}{view} method takes the \mintinline{python}{axes} attribute because it provides the region in graphic coordinates \gbase\ that can be used to query back into data to select a subset as described in \autoref{sec:math:data:sheaf}. We require that the \mintinline{python}{view} method be atomic so that we do not risk race conditions. Atomicity means that the values cannot change after the method is called in draw until a new call to draw\cite{ullmanFirstCourseDatabase2008}, which ensures the integrity of the section.

@@

 \subsection{Scatter, Line, and Bar Artists}
 \label{sec:code:artists}
 \begin{figure}[H]
    \centering 
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=\textwidth]{figures/code/scatter_0.png}
        \caption{}
        \label{fig:code:scatter}
    \end{subfigure}
    \begin{subfigure}{.48\textwidth}
        \includegraphics[width=\textwidth]{figures/code/line_1.png}
        \caption{}
        \label{fig:code:line}
    \end{subfigure}
  \caption{Scatter plot and line plot implemented using \mintinline{python}{Point} and \mintinline{python}{Line} artists and fiber bundle inspired data models. Matplotlib is used for the rendering.}
  \label{fig:code:scatter_line}
\end{figure}
The figure in \autoref{fig:code:scatter} is described by \autoref{eq:math:artist:q:scatter}. This is implemented via a \mintinline{python}{Line} object where the scatter marker shape is fixed as a circle, and the visual fiber components are x and y position and the facecolor and size of the marker. We only show the \mintinline{python}{q_hat} function here because the \mintinline{python}{__init__, draw} are inherited from the prototype artist \mintinline{Python}{ArtistClass}.

The \mintinline{python}{view} method repackages the data as a fiber component indexed table of vertices. Even though the \mintinline{python}{view} is fiber indexed, each vertex at an index \dbasepoint\ has corresponding values in section $\dsection(\dbasepoint_{i})$. This means that all the data on one vertex maps to one glyph.
\begin{minted}{python}
class Point(ArtistClass, mcollections.Collection):
  def q_hat(self, x, y, s, facecolors): #\hat{Q}
    # construct geometries of circle glyphs
    self._paths = [mpath.Path.circle((xi,yi), radius=si) 
                for (xi, yi, si) in zip(x, y, s)] 
    # set attributes of glyphs, these are vectorized 
    # circles and facecolors are lists of the same size
    self.set_facecolors(facecolors)
\end{minted} 
In \mintinline{python}{q_hat}, the \vsection\ components are used to construct the vector path of each circular marker with center \texttt{(x,y)} and size \texttt{x} and set the colors of each circle. This is done via the \mintinline{python}{Path.circle} object. 
\begin{minted}{python}
class Line(ArtistClass, mcollections.LineCollection):
  def q_hat(self, x, y, color): #\hat{Q}
    #assemble line marks as set of segments 
    segments = [np.vstack((vx, vy)).T for vx, vy 
                in zip(x, y)]
    self.set_segments(segments)
    self.set_color(color)
\end{minted}
To generate \autoref{fig:code:line}, the \mintinline{python}{Line} artist \mintinline{python}{view} method returns a table of edges. Each edge consists of (x,y) points sampled along the line defined by the edge and information such as the color of the edge. As with \mintinline{python}{Point}, the data is then converted into visual variables. In \mintinline{python}{q_hat}, described by \autoref{eq:math:artist:q:line}, this visual representation is composed into a set of line segments, where each segment is the array generated by \mintinline{python}{np.vstack((vx, vy))}. Then the colors of each line segment are set. The colors are guaranteed to correspond to the correct segment because of the atomicity constraint on the view. The implementation of line in Matplotlib does not have this functionality because it has no notion of rows and columns of a table, and therefore no notion of a \dsection. Instead, line is assumed to be points along one edge and therefore has only one color, and the user is responsible for aligning the x and y components and colors along the implicit \dbase\ over which they are plotted.

\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_v.png}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_h.png}
    \end{subfigure}
    \caption{Frequency of Penguin types visualized as discrete bars. }
    \label{fig:code_bar_simple}
\end{figure}
The bar charts in figure~\ref{fig:code_bar_simple} are generated with a \mintinline{python}{Bar} artist. The artist has required visual parameters \vfiber\ of (position, length), and an additional parameter \texttt{orientation} which controls whether the bars are arranged vertically or horizontally. This parameter only applies holistically to the graphic and never to individual data parameters, and highlights how the model encourages explicit differentiation between parameters in \vtotal\ and graphic parameters applied directly to \vmarkd.

@@

    def q_hat(self, position, length, floor, width, 
                    facecolors, edgecolors, offset):
        # offset is passed through via assemblers such as multigroup, 
        # not supposed to be directly tagged to position 
        position = position + offset

@@

\subsection{Data Model}
\label{sec:code:data}
The data input into the \mintinline{python}{Artist} will often be a wrapper class around an existing data structure. This wrapper object must specify the fiber components \dfiber\ and connectivity \dbase\ and have a \mintinline{python}{view} method that returns an atomic object that encapsulates \dsection. To support specifying the fiber bundle, we define a \mintinline{python}{FiberBundle} data class\cite{DataclassesDataClasses}

@@

\begin{figure}[htb]
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/linec_1.png}
    \end{subfigure}
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/lined_1.png}
    \end{subfigure}
\caption{Continuous and discontinuous lines as defined via the same data model, and generated with the same \vartisteq\ \mintinline{python}{Line}}
\label{fig:code:multilines}
\end{figure}
The graphics in figure \autoref{fig:code:multilines} are made using the \mintinline{python}{Line} artist and the \mintinline{python}{GraphData} data source where if told that the data is connected, the data source will check for that connectivity by constructing an adjacency matrix. The multicolored line is a connected graph of edges with each edge function evaluated on 100 samples, 
\begin{minted}{python}
GraphData(FB, edges, verticies, num_samples=100, connect=True)
\end{minted}
which is an arbitrary choice made to display a smooth curve. The axes can also be used to choose an appropriate number of samples. In contrast, the stair chart only needs to be evaluated at the edges of the interval 
\begin{minted}{python}
GraphData(FB, edges, verticies, num_samples=2, connect=False)
\end{minted}
such that one advantage of this model is it helps differentiate graphics that have different artists from graphics that have the same artist but make different assumptions about the source data.

\subsection{Case Study: Penguins}
\label{sec:code_case_study}
Building on the \mintinline{python}{Bar} artist in \autoref{sec:code:artists}, we implement grouped bar charts as these do not exist out of the box in the current version of Matplotlib. Instead, grouped bar charts are often achieved via looping over an implementation of bar, which forces the user to keep track which values are mapped to a single visual element and how that is achieved. For this case study, we use the Palmer Penguins dataset\cite{gormanEcologicalSexualDimorphism2014, horstPalmerpenguinsPalmerArchipelago2020}, packaged as a pandas dataframe\cite{nakhaeeMcnakhaeePalmerpenguins2021} since that is a very commonly used Python labeled data structure.

@@

\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_stacked.png}
        \caption{}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_grouped.png}
        \caption{}
    \end{subfigure}
    \caption{Penguin count disaggregated by island and species}
    \label{fig:code_bar_multi}
\end{figure}
The stacked and grouped bar charts in figure~\ref{fig:code_bar_multi} are both composites of \mintinline{python}{Bar} artists such that the difference between \mintinline{python}{StackedBar} and \mintinline{python}{GroupedBar} is specific to the ways in which the \mintinline{python}{Bar} are stitched together. These two artists have identical constructors and \mintinline{python}{draw} methods. As with \mintinline{python}{Bar}, the orientation is set in the constructor. In both these artists, we separate the transforms \mintinline{python}{V}  that are applied to only one component (column) from transforms  \mintinline{python}{MV} applied to multiple components (columns). This convention allows us to, for example, map one column to position, but multiple to length. In effect, we are decomposing \dtotal\ into $\dtotal_1, \cdots, \dtotal_i, \cdots, \dtotal_n$ via specifications in \mintinline{python}{V} rather than by directly taking subsections of the table. This allows us to ensure shared \dbase\ and coherant \dsection.

>>>>

      def draw(self, renderer):
          # returns K, indexed on fiber then key 
          tau = self.E.view(self.axes) 
          # visual channel encoding applied component wise 
          mu = {p: nu(tau(c))
                   for p, (c, nu) in self.V.items()} 
          self.q_hat(**mu)
          # pass configurations off to the renderer
          super().draw(renderer)
  \end{minted}
 The data is fetched in section \dsection\ via a \mintinline{python}{view} method on the data because the input to the artist is a section on \dtotal. The \mintinline{python}{view} method takes the \mintinline{python}{axes} attribute because it provides the region in graphic coordinates \gbase\ that can be used to query back into data to select a subset as described in \autoref{sec:math:data:sheaf}. We require that the \mintinline{python}{view} method be atomic so that we do not risk race conditions. Atomicity means that the values cannot change after the method is called in draw until a new call to draw\cite{ullmanFirstCourseDatabase2008}, which ensures the integrity of the section.

@@

 \subsection{Scatter, Line, and Bar Artists}
 \label{sec:code:artists}
 \begin{figure}[H]
    \centering 
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=\textwidth]{figures/code/scatter_0.png}
        \caption{}
        \label{fig:code:scatter}
    \end{subfigure}
    \begin{subfigure}{.48\textwidth}
        \includegraphics[width=\textwidth]{figures/code/line_1.png}
        \caption{}
        \label{fig:code:line}
    \end{subfigure}
  \caption{Scatter plot and line plot implemented using \mintinline{python}{Point} and \mintinline{python}{Line} artists and fiber bundle inspired data models. Matplotlib is used for the rendering.}
  \label{fig:code:scatter_line}
\end{figure}
The figure in \autoref{fig:code:scatter} is described by \autoref{eq:eq:math:artist:q:scatter}. This is implemented via a \mintinline{python}{Line} object where the scatter marker shape is fixed as a circle, and the visual fiber components are x and y position and the facecolor and size of the marker. We only show the \mintinline{python}{q_hat} function here because the \mintinline{python}{__init__, draw} are inherited from the prototype artist \mintinline{Python}{ArtistClass}.

The \mintinline{python}{view} method repackages the data as a fiber component indexed table of vertices. Even though the \mintinline{python}{view} is fiber indexed, each vertex at an index \dbasepoint\ has corresponding values in section $\dsection(\dbasepoint_{i})$. This means that all the data on one vertex maps to one glyph.
\begin{minted}{python}
class Point(ArtistClass, mcollections.Collection):
  def q_hat(self, x, y, s, facecolors): #\hat{Q}
    # construct geometries of circle glyphs
    self._paths = [mpath.Path.circle((xi,yi), radius=si) 
                for (xi, yi, si) in zip(x, y, s)] 
    # set attributes of glyphs, these are vectorized 
    # circles and facecolors are lists of the same size
    self.set_facecolors(facecolors)
\end{minted} 
In \mintinline{python}{q_hat}, the \vsection\ components are used to construct the vector path of each circular marker with center \texttt{(x,y)} and size \texttt{x} and set the colors of each circle. This is done via the \mintinline{python}{Path.circle} object. 
\begin{minted}{python}
class Line(ArtistClass, mcollections.LineCollection):
  def q_hat(self, x, y, color): #\hat{Q}
    #assemble line marks as set of segments 
    segments = [np.vstack((vx, vy)).T for vx, vy 
                in zip(x, y)]
    self.set_segments(segments)
    self.set_color(color)
\end{minted}
To generate \autoref{fig:code:line}, the \mintinline{python}{Line} artist \mintinline{python}{view} method returns a table of edges. Each edge consists of (x,y) points sampled along the line defined by the edge and information such as the color of the edge. As with \mintinline{python}{Point}, the data is then converted into visual variables. In \mintinline{python}{q_hat}, described by \autoref{eq:eq:math:artist:q:line}, this visual representation is composed into a set of line segments, where each segment is the array generated by \mintinline{python}{np.vstack((vx, vy))}. Then the colors of each line segment are set. The colors are guaranteed to correspond to the correct segment because of the atomicity constraint on the view. The implementation of line in Matplotlib does not have this functionality because it has no notion of rows and columns of a table, and therefore no notion of a \dsection. Instead, line is assumed to be points along one edge and therefore has only one color, and the user is responsible for aligning the x and y components and colors along the implicit \dbase\ over which they are plotted.

\begin{figure}[H]
    \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_v.png}
    \end{subfigure}
    \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_h.png}
    \end{subfigure}
    \caption{Frequency of Penguin types visualized as discrete bars. }
    \label{fig:code_bar_simple}
\end{figure}
The bar charts in figure~\ref{fig:code_bar_simple} are generated with a \mintinline{python}{Bar} artist. The artist has required visual parameters \vfiber\ of (position, length), and an additional parameter \texttt{orientation} which controls whether the bars are arranged vertically or horizontally. This parameter only applies holistically to the graphic and never to individual data parameters, and highlights how the model encourages explicit differentiation between parameters in \vtotal\ and graphic parameters applied directly to \vmarkd.

@@

    def q_hat(self, position, length, floor, width, 
                    facecolors, edgecolors, offset): 
        position = position + offset

@@

\subsection{Data Model}
\label{sec:code:data}
The data input into the \mintinline{python}{Artist} will often be a wrapper class around an existing data structure. This wrapper object must specify the fiber components \dfiber\ and connectivity \dbase\ and have a \mintinline{python}{view} method that returns an atomic object that encapsulates \dsection. To support specifying the fiber bundle, we define a \mintinline{python}{FiberBundle} data class\cite{DataClasses}

@@

\begin{figure}[htb]
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=\textwidth]{figures/code/linec_1.png}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=\textwidth]{figures/code/lined_1.png}
    \end{subfigure}
\caption{Continuous and discontinuous lines as defined via the same data model, and generated with the same \vartisteq\ \mintinline{python}{Line}}
\label{fig:code:multilines}
\end{figure}
The graphics in figure \autoref{fig:code:multilines} are made using the \mintinline{python}{Line} artist and the \mintinline{python}{GraphData} data source where if told that the data is connected, the data source will check for that connectivity by constructing an adjacency matrix. The multicolored line is a connected graph of edges with each edge function evaluated on 100 samples, 
\begin{minted}{python}
GraphData(FB, edges, vertices, num_samples=100, connect=True)
\end{minted}
which is an arbitrary choice made to display a smooth curve. The axes can also be used to choose an appropriate number of samples. In contrast, the stair chart only needs to be evaluated at the edges of the interval 
\begin{minted}{python}
GraphData(FB, edges, vertices, num_samples=2, connect=False)
\end{minted}
such that one advantage of this model is it helps differentiate graphics that have different artists from graphics that have the same artist but make different assumptions about the source data.

\subsection{Case Study: Penguins}
\label{sec:code_case_study}
Building on the \mintinline{python}{Bar} artist in \autoref{sec:code:artists}, we implement grouped bar charts as these do not exist out of the box in the current version of Matplotlib. Instead, grouped bar charts are often achieved via looping over an implementation of bar, which forces the user to keep track which values are mapped to a single visual element and how that is achieved. For this case study, we use the Palmer Penguins dataset\cite{gormanEcologicalSexualDimorphism2014, horstPalmerpenguinsPalmerArchipelago2020}, packaged as a pandas data frame\cite{nakhaeeMcnakhaeePalmerpenguins2021} since that is a very commonly used Python labeled data structure.

@@

\begin{figure}[H]
    \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_stacked.png}
        \caption{}
    \end{subfigure}
    \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_grouped.png}
        \caption{}
    \end{subfigure}
    \caption{Penguin count disaggregated by island and species}
    \label{fig:code_bar_multi}
\end{figure}
The stacked and grouped bar charts in figure~\ref{fig:code_bar_multi} are both composites of \mintinline{python}{Bar} artists such that the difference between \mintinline{python}{StackedBar} and \mintinline{python}{GroupedBar} is specific to the ways in which the \mintinline{python}{Bar} are stitched together. These two artists have identical constructors and \mintinline{python}{draw} methods. As with \mintinline{python}{Bar}, the orientation is set in the constructor. In both these artists, we separate the transforms \mintinline{python}{V}  that are applied to only one component (column) from transforms  \mintinline{python}{MV} applied to multiple components (columns). This convention allows us to, for example, map one column to position, but multiple to length. In effect, we are decomposing \dtotal\ into $\dtotal_1, \cdots, \dtotal_i, \cdots, \dtotal_n$ via specifications in \mintinline{python}{V} rather than by directly taking subsections of the table. This allows us to ensure shared \dbase\ and coherent \dsection.

### 2021-05-09 draft/sections/intro.tex
clean ups

\documentclass[../main.tex]{subfiles}
\begin{document}
\section{Introduction}
\label{sec:intro}
\begin{figure}[h]
    \includegraphics[width=1\textwidth]{figures/intro/functions.png}
    \caption{Building block visualization libraries implement independent reusable functions that map data to visual representations. For example, position encoding functions map the latitude and longitude values in the table (a) to locations on the map (b). These same encoding functions map dates and precipitation (a) to x and y positions in the timeseries plot (c). Color encoding functions map the precipitation (a) to the color of the point in the map (a) and color encoding functions map names of weather stations (a) to colors in both the map (a) and timeseries (b). Functions, which we call \textit{artists}, compose these encoding functions into attributes of visual elements. The \texttt{point} artist transforms the output of the position, facecolor, and edgecolor maps into attributes of each point in the map (a), while the \texttt{line} artist transforms the output of the position and color maps into attributes of each line in the timeseries plot (c).}
    \label{fig:intro:artist}
\end{figure}

Visualizations, by definition, reflect something of the underlying structure and semantics\cite{friendlyBriefHistoryData2008} of the data, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}. We define visualization components to be structure preserving maps from data to visual representations. As illustrated in \autoref{fig:intro:artist}, visualizations map weather station precipitation data (a) into graphical elements such as points (b) and lines (c) and map individual data components (columns) into components of the visualization such as position or color. For example, the gray position encoder function converts the latitude and longitude to x and y positions, and the color encoders map temperature and station name to colors. A \texttt{point} function composites these encodings into attributes of a point in the map (a). The \texttt{line} function composites encoders that convert station name to color and date and precipitation to x and y positions into a piece of a line in the timeseries plot (c). We identify the structure these maps must preserve as the \textit{continuity} of the data and the \textit{equivariance} of the data and visual components.

@@

\subsection{Structure}
\label{sec:intro:structure}
\begin{figure}[H]
        \includegraphics[width=1\textwidth]{figures/intro/equivariant.png}
        \caption{The table (a) lists precipitation in millimeters in blue. These values are converted to  inches, in green, by dividing the millimeter values by 25.4. This conversion is a scaling action, represented by the orange arrows in (a). The \textit{equivariant} action is the scaling in the timeseries (b), which is represented by the orange fill between the green and blue lines. In (c), the scaling is illustrated in the different millimeter and inches y axis labeling of data that appears identical since the distances were scaled by a constant factor that is now incorporated in the labeling.}
      \label{fig:intro:equivariance}
\end{figure}
The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who introduced a notion of structure preservation that we formally describe in terms of \textit{equivariance} and \textit{continuity}. Bertin proposes that there are categories of visual encodings-such as position, shape, color, and texture-that preserve the properties of the measurement type, quantitative or qualitative, of the encoded data. For example, in \autoref{fig:intro:equivariance}, the blue precipitation data in millimeters in the table (a) is converted to inches. This scaling action is represented by the green arrows that translate the precipitation into the scaled values in green. For this visualization to be equivariant, this same scaling factor must be present in the timeseries representation of the data (b). The precipitation in green is an \textit{equivariant} scaling of the precipitation in blue, meaning that the y-values of the green line is $\frac{1}{25.4}$ that of the y-values of the blue precipitation timeseries. The orange fill in the timeseries (b) is the scaling equivalent to the arrows in the table (a). By definition, the distances in the millimeter data were scaled equally \cite{stothersSimilarityGroup} when converted to inches, which means the shapes of the graphs are equivalent when plotted against y-axes that preserve relative distance (c). This visualization is also equivariant to the table (a), but this is indicated through labeling rather than the line plots. The idea of equivariance is formally defined as the mapping of a binary operator from the data domain to the visual domain in Mackinlay's \textit{A Presentation Tool}(APT) model\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomaticDesignGraphical1987}. The algebraic model of visualization proposed by Kindlmann and Scheidegger uses equivariance to refer generally to invertible binary transformations\cite{kindlmannAlgebraicProcessVisualization2014}, which are mathematical groups \cite{shadrachIntroductionGroups2017}. Our model defines \textit{equivariance} in terms of monoid actions, which are a more restrictive set than all binary operations and more general than groups. As with the algebraic model, our model also defines structure preservation as commutative mappings from data space to representation space to graphic space, but our model uses topology to explicitly include continuity.

@@

\subsection{Tools}
\label{sec:intro:data:tools}
\begin{figure}[H]
    \begin{subfigure}{.3\textwidth}
        \includegraphics[width=1\textwidth]{figures/intro/table.png}
        \caption{}
        \label{fig:intro:table}
    \end{subfigure}
    \begin{subfigure}{.3\textwidth}
        \includegraphics[width=1\textwidth]{figures/intro/landsat.png}
        \caption{}
        \label{fig:intro:image}
    \end{subfigure}
    \begin{subfigure}{.3\textwidth}
        \includegraphics[width=1\textwidth]{figures/math/graph.png}
        \caption{}
        \label{fig:intro:graph}
    \end{subfigure}
    \caption{Visualization libraries, especially ones tied to specific domains, tend to be architectured around a core data structure, such as tables \autoref{fig:intro:table}, images \autoref{fig:intro:image}, or networks \autoref{fig:intro:graph}.
    %%revise as combined fig?
    }
    \label{fig:intro:data}
\end{figure}

Most information visualization software design patterns are tuned to specific data structures and domains, as categorized by Heer and Agrawala\cite{HeerSoftware2006}. For users who generally work in one domain--such relational tables (\autoref{fig:intro:table}), images (\autoref{fig:intro:image}), or graphs (\autoref{fig:intro:graph})--well defined data space (and corresponding visual space\cite{chiTaxonomyVisualizationTechniques2000}) often yields a coherent user experience\cite{heerDeclarative2010}. This coherant experience is often not extensible; developers who want to build new visualizations on top of these libraries must work around the existing assumptions, sometimes in ways that break the architecture model of the libraries. For example tools influenced by APT that assume that data is a relational table integrate computation into the visualization pipeline. This is a wide array of tools, including Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}. Since these libraries represent data as a table, and computations on tables are well defined\cite{ullmanFirstCourseDatabase2008}, these libraries implement computation as part of the visualization process. This is also true of tools that support images, such as the biology oriented ImageJ\cite{schneiderNIHImageImageJ2012} and  Napari\cite{nicholas_sofroniew_2021_4533308} or the digital humanities ImageJ macro ImagePlot\cite{studiesCulturevisImageplot2021}. While these tools often have some support for visualizing non image components of the data, the architecture is oriented towards building plugins into the existing system \cite{WritingPlugins} where the image is the core data structure. Tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008} are used to  visualize and manipulate graphs. As with tables and images, extending network libraries to work with other types of data either requires breaking the libraries internal model of how data is structured and what data transformations are allowed or implementing a model for other types of data structures alongside the network model. Our model aims to identify which computations are manipulations of the data and which are necessary for the visual encoding; for example data is often aggregated for a bar chart but it is not required, while a boxplot by definition requires computing distribution statistics\cite{wickham40YearsBoxplots2011}. Disentangling the computation from the visual transforms allows us to determine whether the visualization library should implement computations or if they should be implemented in data space.

>>>>

\documentclass[../main.tex]{subfiles}
\begin{document}
\section{Introduction}
\label{sec:intro}
\begin{figure}[h]
    \includegraphics[width=1\textwidth]{figures/intro/functions.png}
    \caption{Building block visualization libraries implement independent reusable functions that map data to visual representations. For example, position encoding functions map the latitude and longitude values in the table (a) to locations on the map (b). These same encoding functions map dates and precipitation (a) to x and y positions in the time series plot (c). Color encoding functions map the precipitation (a) to the color of the point in the map (a) and color encoding functions map names of weather stations (a) to colors in both the map (a) and time series (b). Functions, which we call \textit{artists}, compose these encoding functions into attributes of visual elements. The \texttt{point} artist transforms the output of the position, face color, and edge color maps into attributes of each point in the map (a), while the \texttt{line} artist transforms the output of the position and color maps into attributes of each line in the time series plot (c).}
    \label{fig:intro:artist}
\end{figure}

Visualizations, by definition, reflect something of the underlying structure and semantics\cite{friendlyBriefHistoryData2008} of the data, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}. We define visualization components to be structure preserving maps from data to visual representations. As illustrated in \autoref{fig:intro:artist}, visualizations map weather station precipitation data (a) into graphical elements such as points (b) and lines (c) and map individual data components (columns) into components of the visualization such as position or color. For example, the gray position encoder function converts the latitude and longitude to x and y positions, and the color encoders map temperature and station name to colors. A \texttt{point} function composites these encodings into attributes of a point in the map (a). The \texttt{line} function composites encoders that convert station name to color and date and precipitation to x and y positions into a piece of a line in the time series plot (c). We identify the structure these maps must preserve as the \textit{continuity} of the data and the \textit{equivariance} of the data and visual components.

@@

\subsection{Structure}
\label{sec:intro:structure}
\begin{figure}[H]
        \includegraphics[width=1\textwidth]{figures/intro/equivariant.png}
        \caption{The table (a) lists precipitation in millimeters in blue. These values are converted to  inches, in green, by dividing the millimeter values by 25.4. This conversion is a scaling action, represented by the orange arrows in (a). The \textit{equivariant} action is the scaling in the time series (b), which is represented by the orange fill between the green and blue lines. In (c), the scaling is illustrated in the different millimeter and inches y axis labeling of data that appears identical since the distances were scaled by a constant factor that is now incorporated in the labeling.}
      \label{fig:intro:equivariance}
\end{figure}
The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who introduced a notion of structure preservation that we formally describe in terms of \textit{equivariance} and \textit{continuity}. Bertin proposes that there are categories of visual encodings-such as position, shape, color, and texture-that preserve the properties of the measurement type, quantitative or qualitative, of the encoded data. For example, in \autoref{fig:intro:equivariance}, the blue precipitation data in millimeters in the table (a) is converted to inches. This scaling action is represented by the green arrows that translate the precipitation into the scaled values in green. For this visualization to be equivariant, this same scaling factor must be present in the time series representation of the data (b). The precipitation in green is an \textit{equivariant} scaling of the precipitation in blue, meaning that the y-values of the green line is $\frac{1}{25.4}$ that of the y-values of the blue precipitation time series. The orange fill in the time series (b) is the scaling equivalent to the arrows in the table (a). By definition, the distances in the millimeter data were scaled equally \cite{stothersKleinViewGeometry} when converted to inches, which means the shapes of the graphs are equivalent when plotted against y-axes that preserve relative distance (c). This visualization is also equivariant to the table (a), but this is indicated through labeling rather than the line plots. The idea of equivariance is formally defined as the mapping of a binary operator from the data domain to the visual domain in Mackinlay's \textit{A Presentation Tool}(APT) model\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomaticDesignGraphical1987}. The algebraic model of visualization proposed by Kindlmann and Scheidegger uses equivariance to refer generally to invertible binary transformations\cite{kindlmannAlgebraicProcessVisualization2014}, which are mathematical groups \cite{shadrachIntroductionGroups2017}. Our model defines \textit{equivariance} in terms of monoid actions, which are a more restrictive set than all binary operations and more general than groups. As with the algebraic model, our model also defines structure preservation as commutative mappings from data space to representation space to graphic space, but our model uses topology to explicitly include continuity.

@@

\subsection{Tools}
\label{sec:intro:data:tools}
\begin{figure}[H]
    \begin{subfigure}{.3\textwidth}
        \includegraphics[width=1\textwidth]{figures/intro/table.png}
        \caption{}
        \label{fig:intro:table}
    \end{subfigure}
    \begin{subfigure}{.3\textwidth}
        \includegraphics[width=1\textwidth]{figures/intro/landsat.png}
        \caption{}
        \label{fig:intro:image}
    \end{subfigure}
    \begin{subfigure}{.3\textwidth}
        \includegraphics[width=1\textwidth]{figures/math/graph.png}
        \caption{}
        \label{fig:intro:graph}
    \end{subfigure}
    \caption{Visualization libraries, especially ones tied to specific domains, often are designed around a core data structure, such as tables \autoref{fig:intro:table}, images \autoref{fig:intro:image}, or networks \autoref{fig:intro:graph}.
    %%revise as combined fig?
    }
    \label{fig:intro:data}
\end{figure}

Most information visualization software design patterns are tuned to specific data structures and domains, as categorized by Heer and Agrawala\cite{HeerSoftware2006}. For users who generally work in one domain--such relational tables (\autoref{fig:intro:table}), images (\autoref{fig:intro:image}), or graphs (\autoref{fig:intro:graph})--well defined data space (and corresponding visual space\cite{chiTaxonomyVisualizationTechniques2000}) often yields a coherent user experience\cite{heerDeclarative2010}. This coherent experience is often not extensible; developers who want to build new visualizations on top of these libraries must work around the existing assumptions, sometimes in ways that break the architecture model of the libraries. For example tools influenced by APT that assume that data is a relational table integrate computation into the visualization pipeline. This is a wide array of tools, including Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}. Since these libraries represent data as a table, and computations on tables are well defined\cite{ullmanFirstCourseDatabase2008}, these libraries implement computation as part of the visualization process. This is also true of tools that support images, such as the biology oriented ImageJ\cite{schneiderNIHImageImageJ2012} and  Napari\cite{nicholas_sofroniew_2021_4533308} or the digital humanities ImageJ macro ImagePlot\cite{studiesCulturevisImageplot2021}. While these tools often have some support for visualizing non image components of the data, the architecture is oriented towards building plugins into the existing system \cite{WritingPlugins} where the image is the core data structure. Tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008} are used to  visualize and manipulate graphs. As with tables and images, extending network libraries to work with other types of data either requires breaking the libraries internal model of how data is structured and what data transformations are allowed or implementing a model for other types of data structures alongside the network model. Our model aims to identify which computations are manipulations of the data and which are necessary for the visual encoding; for example data is often aggregated for a bar chart but it is not required, while a boxplot by definition requires computing distribution statistics\cite{wickham40YearsBoxplots2011}. Disentangling the computation from the visual transforms allows us to determine whether the visualization library should implement computations or if they should be implemented in data space.
